Embedding Dimension of Polyhedral Losses

### 2020-01-31 colt-20/embed-dim.tex
section 4 is under seige

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Higher dimensions}\label{sec:d-dim}
Our newfound understanding of the 1-dimensional embedding characterization leads us to realize that a large class of properties are not $1$-embeddable.
\cite{finocchiaro2019embedding} have shown a trivial upper bound that every property is $(n-1)$-embeddable\jessie{already mentioned earlier}, but it is an open question as to how tight we can get the bounds on embedding dimension.
Reducing embedding dimension significantly below $(n-1)$ can help provide a computational speedup in the optimization problem, which is a function of the surrogate loss dimension.
This is especially important as the number of possible outcomes $n$ grows large.

Below, we generalize our characterization to $d$-embeddability for $d \geq 2$.
Having a finite number of embedding points and restricting to polyhedral functions allows us to study the subgradient sets of the loss at embedding points, which we formalize in Section~\ref{subsec:sub-sets}.
Using a general theory of polytopes, we can then present our general $d$-dimensional characterization in terms of optimality and monotonicity conditions.
In Section~\ref{subsec:opt-conditions}, we present a quadratic feasibility program in order to understand necessary conditions for $d$-embeddability by satisfying optimality conditions, yielding new lower bounds for embedding dimension in Corollary~\ref{cor:d-embeddable-char}.

@@

Recall that if $\ell: \R \to \nonnegreals^{\Y}$ is embedded by $L: \reals^d \to \nonnegreals^{\Y}$, then each $r \in \R$ is embedded at some point $u_r \in \reals^d$.
In particular, $u_r$ must minimize $\inprod{p}{L(u)}$ if and only if $r$ minimizes $\inprod{p}{\ell(r')}$.

The key to our approach is to study as first-class objects the sets of all subgradients\footnote{Recall that a subgradient of e.g. the convex function $L(\cdot)_y: \reals^d \to \reals$ at a point $u$ is a vector $v \in \reals^d$ such that $L(u')_y \geq L(u)_y + \inprod{v}{u'-u}$ for all $u'$.} of $L$ at these embedding points.
To settle the question of whether a consistent surrogate exists in $d$ dimensions, it surprisingly turns out to be necessary and sufficient to consider conditions on these sets alone.
In particular, we use heavily that a convex function is minimized at a point if and only if the origin $\vec{0}$ is in its subgradient set at that point.

Therefore, we abstract to consider collections of sets $T^r_y$, which aspire to be the subgradient sets of a calibrated polyhedral surrogate $L(\cdot)_y$ at $u_r$.
Throughout, we fix $r$ and remove it from the subscript for ease of exposition.
Note that if $L(\cdot)_y$ is a polyhedral function on $\reals^d$, then all of its subgradient sets are (bounded) closed polytopes.\bo{cite}\jessie{\cite{rockafellar1997convex} ? in my notes, but I don't have the exact theorem number.}

\begin{definition}[$\T$, $D(\T)$]
  We use the following notation for a collection of closed polytopes, with implicit parameters $d$ and $r$: $\T = \{T_y \subseteq \reals^d : y \in \Y\}$.

  Given a distribution $p \in \simplex$, we write the $p$-weighted Minkowski sum of $\T$ as
    \[ \oplus_p \T := \left\{ \sum_{y \in \Y} p_y x_y ~\Big|~ x_1 \in T_1, \dots, x_y \in T_y \right\} , \]
  in other words, the Minkowski sum of the scaled sets $\{p_y T_y : y \in \Y\}$.

  Finally, we associate with $\T$ a set of distributions $D(\T) = \{ p \in \simplex : \vec 0 \in \oplus_p \T\}$.
\end{definition}

The importance of the $p$-weighted Minkowski sum and of $D(\T)$ are that they capture the distributions $p$ for which $u_r$ minimizes expected loss, assuming that $\T$ correspond to some polyhedral $L$.
\begin{proposition} \label{prop:minkowski-0-opt}
  If $T_y$ is the subgradient set of each $L(u_r)_y$, then $D(\T)$ is the set of distributions $p \in \simplex$ for which the embedding point $u_r$ minimizes expected $L$-loss.
\end{proposition}
\begin{proof}
  Any convex function $f$ is minimized at $u$ if and only if $\vec 0$ is a subgradient of $f$ at $u$.
  Let $f(u) = \inprod{p}{L(u)}$.
  Then the subgradient set of $f$ at $u$ is $\oplus_p \T$.
  This follows from the basic properties that if $f_1,f_2$ are convex with subgradient sets $T_1,T_2$ at $u$, then $\alpha f_1$ has subgradient set $\alpha T_1$ and $f_1 + f_2$ has subgradient set $T_1 \oplus T_2$, the Minkowski sum.
  So $\vec 0$ is a subgradient of $\inprod{p}{L(u_r)}$ if and only if $\vec 0 \in \oplus_p \T$, which by definition occurs if and only if $p \in D(\T)$.
\end{proof}
This fact will be vital for characterizing when $\ell$ is correctly embedded by some $L$ whose subgradient sets are $\T$ for each $r \in \R$.
We leverage it next.

\subsection{General characterization}

We now give a general characterization of when a discrete loss $\ell$ can be embedded into $d$ dimensions, i.e. when a consistent polyhedral surrogate $L: \reals^d \to \reals^{\Y}_{\geq 0}$ exists.

Two conditions are required: \emph{optimality} and \emph{monotonicity}.
Optimality enforces that the surrogate is minimized precisely when and where it should be.
It says that for each discrete prediction $r$ and set of distributions $\gamma_r$ for which it is optimal, there exists a collection of polytopes $\T$ (presumably the subgradient sets of $L(u_r)_y$ for each $y$) realizing Proposition~\ref{prop:minkowski-0-opt}.
Monotonicity says that these individual polytopes can actually be glued together to form the subgradients of some convex loss function $L$.

@@

We now investigate the optimality condition of Theorem \ref{thm:general-char}, for two purposes.
First, we aim to greatly narrow the search space for constructing low-dimensional surrogate loss functions for a given discrete loss.
The tools we construct in this section aid in this task by constraining or constructing feasible subgradient sets $\T$ given a level set $\gamma_r$.
Second, we wish to prove impossibilities, i.e. lower bounds on the embedding dimension of a given discrete loss (an apparently hard problem).

By dropping monotonicity from Theorem \ref{thm:general-char}, we obtain $|\R|$ independent optimality conditions, $D(\T) = \gamma_r$, that must be satisfied in order to have a $d$-dimensional embedding.
For upper bounds, one must construct a $\T$ satisfying $D(\T) = \gamma_r$ for all $r$ (and then turn to satisfying monotonicity).
For lower bounds, it suffices to prove nonexistence of any such $\T$, for any $r$.

To accomplish these, we construct a quadratic feasibility program (QFP, Definition \ref{def:qfp}) that takes as input a polytope $C \subseteq \simplex$ and a parameter $d \in \mathbb{N}$ and (roughly) attempts to construct $\T$.
We will prove the following main results.
\begin{theorem} \label{thm:opt-iff-qfp}
  Given a convex polytope $C \subseteq \simplex$, there exist polytopes $\T$ in $\reals^d$ such that $D(\T) = C$ if and only if there is a feasible solution to the QFP (Definition \ref{def:qfp}) with parameter $d$.
\end{theorem}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Higher dimensions}\label{sec:d-dim}
Our characteriation of 1-embeddable losses reveals a large class of properties are not $1$-embeddable.
On the other hand, previous work has shown that every discrete loss is $(n-1)$-embeddable
\citep{finocchiaro2019embedding}.
A natural next question, therefore, is to characterize $d$-embeddable losses for $2 \leq d \leq n-2$, or failing that, to develop tighter bounds on the embedding dimension of various losses.
An embedding dimension significantly below $(n-1)$, such as $O(\log n)$ as with abstain loss, can lead to faster downstream optimization, as rates and computation times can both depend on the surrogate loss dimension.
This dependence is especially important as the number of possible outcomes $n$ grows large.

In this section, we develop a characterization of $d$-embeddable discrete losses for $d \geq 2$.
We begin with some basic facts and definitions about polytopes and their Minkowski sums, which naturally arise when considering the subgradients of a polyhedral surrogate loss (\S~\ref{subsec:sub-sets}).
From these definitions, we can state a somewhat immediate characterization of $d$-embeddable losses in terms of polytopes that satisfy certain optimality and monotonicity conditions (\S~\ref{subsec:gener-char}).
We then explore the optimality condition further, and through facts about Minkowski sums, slowly remove mentions of polytopes from the condition until we arrive at a quadratic feasibility program to test whether such polytopes exist (\S~\ref{subsec:opt-conditions}).
Finally, we discuss the proof of correctness of our feasibility program together with intuition.

\raf{Orphaned; add back:} necessary conditions for $d$-embeddability by satisfying optimality conditions, yielding new lower bounds for embedding dimension in Corollary~\ref{cor:d-embeddable-char}.

@@

Recall that if $\ell: \R \to \nonnegreals^{\Y}$ is embedded by $L: \reals^d \to \nonnegreals^{\Y}$, then each $r \in \R$ is embedded at some point $\varphi(r) \in \reals^d$.
In particular, $\varphi(r)$ must minimize $\inprod{p}{L(u)}$ if and only if $r$ minimizes $\inprod{p}{\ell(r')}$.
The key to our approach is to study as first-class objects the sets of all subgradients\footnote{Recall that a subgradient of e.g. the convex function $L(\cdot)_y: \reals^d \to \reals$ at a point $u$ is a vector $v \in \reals^d$ such that $L(u')_y \geq L(u)_y + \inprod{v}{u'-u}$ for all $u'$.} of $L$ at these embedding points.
The question of whether a calibrated polyhedral surrogate exists in $d$ dimensions essentially reduces to conditions on these sets alone.
\raft{The ``essentially'' is because we need loss values to match.  It's not clear if that comes for free, i.e., if you embed the correct property, can you always modify the loss to match $\ell$?}
In particular, we use the fact that a convex function is minimized at $u$ if and only if $\vec{0}$ is in its subgradient set at $u$.

Therefore, we consider collections of sets $T^r_y$, which intuitively aspire to be the subgradient sets of a calibrated polyhedral surrogate $L(\cdot)_y$ at $\varphi(r)$, denoted $\partial L(\varphi(r))_y$.
Throughout, we often take $r$ as implicit and suppress it from our notation for ease of exposition.
Note that if $L(\cdot)_y$ is a polyhedral function on $\reals^d$, then all of its subgradient sets are (bounded) closed polytopes.\bo{cite}\jessie{\cite{rockafellar1997convex} ? in my notes, but I don't have the exact theorem number.}

\begin{definition}[$\T$, $D(\T)$]
  We write $\T = \{T_y \subseteq \reals^d : y \in \Y\}$ to denote a collection of closed polytopes, with implicit parameter $d$.
  Given a distribution $p \in \simplex$, we write the $p$-weighted Minkowski sum of $\T$ as
  \begin{equation*}
    \oplus_p \T := \mathop{\bigoplus}\limits_{y\in\Y} p_yT_y = \left\{ \sum_{y \in \Y} p_y x_y ~\Big|~ x_y \in T_y \; \forall y\in\Y \right\}~,
  \end{equation*}
  or in other words, the Minkowski sum of the scaled sets $\{p_y T_y : y \in \Y\}$.
  Finally, we associate with $\T$ a set of distributions $D(\T) = \{ p \in \simplex : \vec 0 \in \oplus_p \T\}$.
\end{definition}

The importance of the $p$-weighted Minkowski sum and of $D(\T)$ are that they capture the distributions $p$ for which a point $u$ minimizes expected loss, whenever $\T$ corresponds to the subgradient sets of some polyhedral $L$ at $u$.
In other words, under these conditions, we have $D(\T) = \Gamma_u$, the level set for $u$ of the property $\Gamma$ elicited by $L$.
\begin{lemma} \label{lem:minkowski-0-opt}
  Let $L: \reals^d \to \nonnegreals^{\Y}$ be a polyhedral loss eliciting a property $\Gamma$.
  If for all $y\in\Y$ we have $T_y = \partial L(u)_y$ at some point $u\in\reals^d$, then $D(\T) = \Gamma_u$.
\end{lemma}
\begin{proof}
  Recall that a convex function $f$ is minimized at $u$ if and only if $\vec 0 \in \partial f(u)$.
  % Let $f(u) = \inprod{p}{L(u)}$.
  We thus have $p\in\Gamma_u \iff u \in \argmin_{u'} \inprod{p}{L(u')} \iff \vec 0 \in \partial\inprod{p}{L(u)} = \mathop{\bigoplus}_{y\in\Y} p_y \partial L(u)_y = \oplus_p \T \iff p\in D(\T)$.
  Here we used the basic fact that if $f_1,f_2$ are convex with subgradient sets $T_1,T_2$ at $u$, then $\alpha f_1 + \beta f_2$ has subgradient set $\alpha T_1 \oplus \beta T_2$, the Minkowski sum of the scaled sets.
  % So $\vec 0$ is a subgradient of $\inprod{p}{L(\varphi(r))}$ if and only if $\vec 0 \in \oplus_p \T$, which by definition occurs if and only if $p \in D(\T)$.
\end{proof}
This fact will be vital for characterizing when $\ell$ is correctly embedded by some $L$ whose subgradient sets are $\T^r$ for each $r \in \R$.

\subsection{General characterization}
\label{subsec:gener-char}

We now give a general characterization of when a discrete loss $\ell$ can be embedded into $d$ dimensions, i.e. when a consistent polyhedral surrogate $L: \reals^d \to \reals^{\Y}_{\geq 0}$ exists.
Two conditions are required: \emph{optimality} and \emph{monotonicity}.
Optimality enforces that the surrogate is minimized precisely when and where it should be.
It says that for each discrete prediction $r$ and set of distributions $\gamma_r$ for which it is $\ell$-optimal, there exists a collection of polytopes $\T^r$ such that, were they the subgradients of some polyhedral surrogate $L$ at some point $\varphi(r)$, then $\varphi(r)$ would be $L$-optimal at the same set of distributions $\gamma_r$; more succinctly in light of Lemma~\ref{lem:minkowski-0-opt}, we require $D(\T^r)=\gamma_r$.
Monotonicity says that these individual polytopes can indeed be glued together to form the subgradients of some convex loss function $L$.

@@

We now focus entirely on the optimality condition of Theorem \ref{thm:general-char}, for two purposes.
First, we aim to greatly narrow the search space for constructing low-dimensional surrogate loss functions for a given discrete loss.
The tools we construct in this section aid in this task by constraining or constructing feasible subgradient sets $\T$ given a level set $\gamma_r$.
Second, we wish to prove impossibilities, i.e., lower bounds on the embedding dimension of a given discrete loss (an apparently hard problem).
For such lower bounds, it suffices to drop monotonicity from Theorem \ref{thm:general-char}, leaving us with an independent optimality condition for each $r\in\R$, and show that for any one $r\in\R$, we could not have $d$-dimensional polytopes $\T$ satisfying $D(\T) = \gamma_r$.
% (For upper bounds, one must additionally show monotonicity.)

At first glance, the optimality condition seems difficult to operationalize, as it involves the existence of polytopes, and even if said polytopes are given, it is unclear how to test whether $D(\T)=\gamma_r$.
To begin, consider the latter problem, of understanding the set $D(\T)$ in terms of descriptions of $\T$, and in particular, of writing conditions on $\T$ such that $D(\T)$ is equal to a given polytope $C\subseteq\simplex$.
We now that, by writing $C$ in its halfspace and vertex representations, respectively, we can give two such conditions.

\begin{condition}[Halfspace condition]\label{cond:H-condition}
	A collection of polytopes $\T$ and a polytope $C\subseteq\simplex$ defined by $C = \{p \in \simplex : Bp \geq \vec 0\}$ \emph{satisfy the halfspace condition} if there exist $v_1, \ldots, v_k \in \reals^d$ such that, for all $i \in [k]$ and $y \in \Y$, for all $x \in T_y$, we have $\inprod{v_i}{x} \leq B_{iy}$.
\end{condition}
\begin{condition}[Vertex condition]\label{cond:V-condition}
	A collection of polytopes $\T$ and a polytope $C\subseteq\simplex$ defined by $C = \conv(\{p^1, \ldots, p^\ell\})$ \emph{satisfy the vertex condition} if for all $j \in [\ell]$, $0 \in \oplus_{p^j} \T$. %% Equivalent: there exist $\{x_{jy} \in T_y : j \in [\ell]\}$ such that, for all $j \in [\ell]$, $\sum_{y \in \Y} p^j_y x_{jy} = \vec 0$.
\end{condition}

\begin{theorem} \label{thm:vertex-halfspace-opt}
  Let the polytopes $\T = \{T_y \subseteq \reals^d : y \in \Y\}$ and $C$ be given, with $C = \conv(\{p^1,\ldots,p^{\ell}\}) = \{p: Bp \geq \vec 0\}$ for $B \in \reals^{k \times n}$.
  We have $D(\T) = C$ if and only if both the halfspace and vertex conditions hold.
\end{theorem}

The two conditions above give us a much better understanding of when a given set of polytopes $\T$ satisfies the optimality condition.
We are still left with the problem, however, of understanding when such a set $\T$ exists.
Intuitively, the biggest hurdle that remains is the quantification over sets of polytopes, a massive search space.
Surprisingly, one can reduce this search to a simple quadratic feasibility program, which we now give.

Our program will consist of variables

are a set of normal vectors $\{v_i \in \reals^d : i \in [k]\}$ and a set of vertices $\{x^j_y \in \reals^d : j \in [\ell], y \in \Y\}$.
These correspond to a (possibly relaxed) halfspace representation and (possibly incomplete) vertex representation of the subgradient polytopes $\T$.

\section{OLD / TO MOVE}

To accomplish our goal, we construct a quadratic feasibility program (QFP, Definition \ref{def:qfp}) that takes as input a polytope $C \subseteq \simplex$ and a parameter $d \in \mathbb{N}$ and (roughly) attempts to construct $\T$.
We will prove the following main results.
\begin{theorem} \label{thm:opt-iff-qfp}
  Given a convex polytope $C \subseteq \simplex$, there exist polytopes $\T$ in $\reals^d$ such that $D(\T) = C$ if and only if there is a feasible solution to the QFP (Definition \ref{def:qfp}) with parameter $d$.
\end{theorem}
